Gravitational wave sources in our Galactic backyard: Predictions for BH and NS binaries in LISA

### 2021-07-23 paper/tex/sections/appendix.tex
round 2 of feedback from Stephen

\section{Assessing the impact of Milky Way model choices}\label{sec:mw_changes}
The model that we use for the Milky Way adds several layers of complexity, accounting for the inside-out growth of the thin disc, using empirically informed star formation histories that are a function of time and assigning metallicities based on the position and age of binaries. In this section, we repeat our main analysis but instead apply a simpler model for the Milky Way in order to assess the effect of these added features. For this purpose, we use model for the Milky Way used in \citet{Breivik+2020} as this is representative of the models used in most previous works.

>>>>

\section{Assessing the impact of Milky Way model choices}\label{app:mw_changes}
The model that we use for the Milky Way adds several layers of complexity, accounting for the inside-out growth of the thin disc, using empirically informed star formation histories that are a function of time and assigning metallicities based on the position and age of binaries. In this section, we repeat our main analysis but instead apply a simpler model for the Milky Way in order to assess the effect of these added features. For this purpose, we use model for the Milky Way used in \citet{Breivik+2020} as this is representative of the models used in most previous works.

### 2021-07-23 paper/tex/sections/comparisons.tex
round 2 of feedback from Stephen

\paragraph{\citet{Lamberts+2018}} presented a new approach to the problem by using the FIRE simulation \citep{Hopkins+2014} to distribute their sources rather than an analytical model of the Milky Way, thus being the first paper in this area to incorporate metallicity dependence into their Milky Way model. \citet{Sesana+2020} followed up on this paper using the same simulated BHBH population and presented updated results for the number of expected BHBH detections. They find significantly fewer BHBHs than our fiducial model despite using the same SNR threshold and LISA mission length.
%
The discrepancy between the results of \citet{Sesana+2020} and those presented in this work could be caused by different treatments of eccentricity. Unlike our work, \citet{Sesana+2020} assume that all binaries are circular for the purpose of detection in LISA, which could result in a lower number of detections by missing eccentric binaries that appear as weaker signals when assumed to be circular. This is especially relevant as we find that around $\BHBHNotCirc{}$ of LISA detectable BHBHs are not circular and around $\BHBHHighlyEccentric{}$ have significant eccentricity (see Section~\ref{sec:fiducial_distributions}). We also improve upon this work by using a larger number of metallicity bins compared to \citet{Sesana+2020}, since a low number of metallicity bins can produce artificial features in the mass distribution of DCOs and possibly affect the detection rate (see Section~\ref{sec:mw_changes}). Finally, it could be that different implicit assumptions in their population synthesis code lead to differences in our results \citep{Toonen+2014}.

\paragraph{\citet{Lau+2020}} focussed on the number of Galactic NSNS binaries that could be detected by LISA. Their study uses the same population synthesis code, COMPAS, as this work, though an earlier version. Despite this, their study finds a much larger number of detections. They make several different physical assumptions in their population synthesis, using the \citet{Fryer+2012} \textit{rapid} remnant mass prescription, limiting the maximum neutron star mass to $2 \unit{M_{\odot}}$ and not implementing PISN. However, we note that none of these assumptions strongly affect the NSNS LISA detection rate (see bottom panel of Fig.~\ref{fig:detection_rates}, models \modRapid{}, \modNSLow{} and \modNoPISN{}) and so this is unlikely to entirely account for our differences. It is also important to highlight that COMPAS has received several improvements and bug fixes since \citet{Vigna-Gomez+2018} (which contains the simulations used by \citet{Lau+2020}) and these could possibly have affected the formation rate of NSNSs.
%
Yet it is most likely that the remaining difference between our results is due to way in which we simulate the Milky Way. \citet{Lau+2020} use a model for the Milky Way similar to that of \citet{Breivik+2020}, which we used to estimate how impactful the choice of MW model was in Appendix~\ref{sec:mw_changes}. We find that this simpler model for the Milky Way could result in an overestimation of the NSNS detection rate by at least a factor of two and so this may explain the discrepancy between our results. \citet{Lau+2020} additionally only uses a single metallicity rather than the two used in \citet{Breivik+2020} and so this effect could be even stronger as their results will have no contribution from lower metallicity systems.

\paragraph{\citet{Breivik+2020}} introduced the population synthesis code COSMIC and presented detections for many different DCO types in LISA using this code. They find that LISA will detect 93 BHBHs, 33 BHNSs and 8 NSNSs in the Milky Way over a 4 year mission. \citet{Breivik+2020} make many different physical assumptions from us, the most notable being that they assume the optimistic CE scenario and that case BB mass transfer is always unstable, whilst also using a simpler model for the Milky Way (see Appendix~\ref{sec:mw_changes}). Thus for better comparison we ran our simulation using model \modCaseBBOpt{} and the Milky Way model from \citet{Breivik+2020}. This results in 97, 101 and 43 detections for BHBHs, BHNSs and NSNSs respectively. Therefore, though we are in very good agreement for BHBHs, we predict much higher rates for BHNSs and NSNSs. These differences are likely due to using a different population synthesis code (COSMIC), which has different underlying physics assumptions from COMPAS. Given our strong agreement for BHBHs, it is possible that COSMIC and COMPAS handle NSs differently and thus lead to different detection rates for DCOs containing NSs. However checking this would require a more in-depth study of the intrinsic formation rate of DCOs containing NSs in the two codes.

\paragraph{\citet{Shao+2021}} most recently investigated the detectability binaries containing BHs in LISA using BSE and a relatively simple model for the Milky Way (assuming a uniform flat disc, constant star formation and a single metallicity). They assume that kicks for NSs formed through ECSN are slightly higher than our work ($50 \unit{km}{s^{-1}}$ instead of $30 \unit{km}{s^{-1}}$). This may account for their particularly low BHNS rate (as the binaries would be more likely to disrupt), which is a factor of 20 lower than ours, but we expect their assumption of the optimistic CE scenario, reduced Wolf-Rayet winds and lower SNR detection threshold would offset this. As we show in Appendix~\ref{sec:mw_changes}, their use of a simpler Milky Way model, especially with only a single metallicity, would lead to an underestimate of the BHBH and BHNS rates, which may explain the discrepancy in our results.

>>>>

\paragraph{\citet{Lamberts+2018}} presented a new approach to the problem by using the FIRE simulation \citep{Hopkins+2014} to distribute their sources rather than an analytical model of the Milky Way, thus being the first paper in this area to incorporate metallicity dependence into their Milky Way model. \citet{Sesana+2020} followed up on this paper using the same simulated BHBH population and presented updated results for the number of expected BHBH detections. They find significantly fewer BHBHs than our fiducial model despite using the same SNR threshold and LISA mission length.
%
The discrepancy between the results of \citet{Sesana+2020} and those presented in this work could be caused by different treatments of eccentricity. Unlike our work, \citet{Sesana+2020} assume that all binaries are circular for the purpose of detection in LISA, which could result in a lower number of detections by missing eccentric binaries that appear as weaker signals when assumed to be circular. This is especially relevant as we find that around $\BHBHNotCirc{}$ of LISA detectable BHBHs are not circular and around $\BHBHHighlyEccentric{}$ have significant eccentricity (see Section~\ref{sec:fiducial_distributions}). We also improve upon this work by using a larger number of metallicity bins compared to \citet{Sesana+2020}, since a low number of metallicity bins can produce artificial features in the mass distribution of DCOs and possibly affect the detection rate (see Appendix~\ref{app:mw_changes}). Finally, it could be that different implicit assumptions in their population synthesis code lead to differences in our results \citep{Toonen+2014}.

\paragraph{\citet{Lau+2020}} focussed on the number of Galactic NSNS binaries that could be detected by LISA. Their study uses the same population synthesis code, COMPAS, as this work, though an earlier version. Despite this, their study finds a much larger number of detections. They make several different physical assumptions in their population synthesis, using the \citet{Fryer+2012} \textit{rapid} remnant mass prescription, limiting the maximum neutron star mass to $2 \unit{M_{\odot}}$ and not implementing PISN. However, we note that none of these assumptions strongly affect the NSNS LISA detection rate (see bottom panel of Fig.~\ref{fig:detection_rates}, models \modRapid{}, \modNSLow{} and \modNoPISN{}) and so this is unlikely to entirely account for our differences. It is also important to highlight that COMPAS has received several improvements and bug fixes since \citet{Vigna-Gomez+2018} (which contains the simulations used by \citet{Lau+2020}) and these could possibly have affected the formation rate of NSNSs.
%
Yet it is most likely that the remaining difference between our results is due to way in which we simulate the Milky Way. \citet{Lau+2020} use a model for the Milky Way similar to that of \citet{Breivik+2020}, which we used to estimate how impactful the choice of MW model was in Appendix~\ref{app:mw_changes}. We find that this simpler model for the Milky Way could result in an overestimation of the NSNS detection rate by at least a factor of two and so this may explain the discrepancy between our results. \citet{Lau+2020} additionally only uses a single metallicity rather than the two used in \citet{Breivik+2020} and so this effect could be even stronger as their results will have no contribution from lower metallicity systems.

\paragraph{\citet{Breivik+2020}} introduced the population synthesis code COSMIC and presented detections for many different DCO types in LISA using this code. They find that LISA will detect 93 BHBHs, 33 BHNSs and 8 NSNSs in the Milky Way over a 4 year mission. \citet{Breivik+2020} make many different physical assumptions from us, the most notable being that they assume the optimistic CE scenario and that case BB mass transfer is always unstable, whilst also using a simpler model for the Milky Way (see Appendix~\ref{app:mw_changes}). Thus for better comparison we ran our simulation using model \modCaseBBOpt{} and the Milky Way model from \citet{Breivik+2020}. This results in 97, 101 and 43 detections for BHBHs, BHNSs and NSNSs respectively. Therefore, though we are in very good agreement for BHBHs, we predict much higher rates for BHNSs and NSNSs. These differences are likely due to using a different population synthesis code (COSMIC), which has different underlying physics assumptions from COMPAS. Given our strong agreement for BHBHs, it is possible that COSMIC and COMPAS handle NSs differently and thus lead to different detection rates for DCOs containing NSs. However checking this would require a more in-depth study of the intrinsic formation rate of DCOs containing NSs in the two codes.

\paragraph{\citet{Shao+2021}} most recently investigated the detectability binaries containing BHs in LISA using BSE and a relatively simple model for the Milky Way (assuming a uniform flat disc, constant star formation and a single metallicity). They assume that kicks for NSs formed through ECSN are slightly higher than our work ($50 \unit{km}{s^{-1}}$ instead of $30 \unit{km}{s^{-1}}$). This may account for their particularly low BHNS rate (as the binaries would be more likely to disrupt), which is a factor of 20 lower than ours, but we expect their assumption of the optimistic CE scenario, reduced Wolf-Rayet winds and lower SNR detection threshold would offset this. As we show in Appendix~\ref{app:mw_changes}, their use of a simpler Milky Way model, especially with only a single metallicity, would lead to an underestimate of the BHBH and BHNS rates, which may explain the discrepancy in our results.

### 2021-07-23 paper/tex/sections/discussion.tex
round 2 of feedback from Stephen

Another discriminator between WDWDs and other DCOs is the eccentricity. WDWDs formed in the disc are thought to be formed through isolated binary formation and have little to no eccentricity \citep[e.g.][]{Nelemans+2001}. Therefore, if any system is detected with anything other than one detectable harmonic, this implies that the system is unlikely to be a WDWD. We find that for a 4(10)-year LISA mission, \BHBHMultipleHarmonicsFourPerc{}(\BHBHMultipleHarmonicsTenPerc{})\% of BHBHs, \BHNSMultipleHarmonicsFourPerc{}(\BHNSMultipleHarmonicsTenPerc{})\% of BHNSs and \NSNSMultipleHarmonicsFourPerc(\NSNSMultipleHarmonicsTenPerc{})\% of NSNSs are detected with multiple harmonics. Both the absolute percentage and the relative improvement with an extended LISA mission is lower for the BHNSs with respect to other DCOs as we find that these BHNSs are less eccentric on average (see Fig.~\ref{fig:fiducial_pdf_distributions} and discussion in Sec. \ref{sec:fiducial_distributions}).

@@

For separating the BHBH and BHNS population one could consider the range of mass ratios that could result in the chirp mass and assume that those that are likely close to unity must be BHBHs. Another possible solution would be the existence of electromagnetic counterparts to the gravitational wave signal. In Section~\ref{sec:pulsar_matching} we consider the possibility of detecting a pulsar within a BHNS or NSNS system. This could be used to identify the type of the source, however it is unlikely that a large fraction of the population will contain pulsars that are beaming towards the Earth.

One could also consider using the eccentricity or orbital frequency to separate the populations since the distributions are different for each DCO type (see Fig.~\ref{fig:fiducial_pdf_distributions}). This method would also pose a challenge as it would likely only rule out some DCO types as the source of the signal rather than provide strong evidence of the exact type.

\subsection{Matching LISA detections to pulsars with SKA}\label{sec:pulsar_matching}
Since the vast majority of the LISA detectable population of DCOs will not merge for many years, the main form of electromagnetic counterpart for the this population is pulsars. Therefore, for this section we focus only on BHNSs and NSNSs since no BHBH system will contain a pulsar. The joint detection of a binary pulsar with LISA and SKA would not only help to constrain the parameters of the binary, but also enable investigation of other compact object physics. A pulsar(PSR)+BH can be provide stringent tests of theories of gravity, in particular the ``No-hair theorem'' \citep{Keane+2015}. Alternatively, an ultrarelativstic PSR+NS system could be used to measure the neutron star equation of state up to an order of magnitude more accurately than other proposed observational constraints \citep{Kyutoku+2019, Thrane+2020}.

We estimate on average, a LISA DCO localised with an angular resolution of $\sigma_{\theta} < 1.3^\circ$ or $0.7^\circ$ (for SKA-1 and SKA-2) will never contain more than one pulsar in its sky localisation region (see Appendix~\ref{app:ska_area}). Given these estimates, and by considering the last panel of Fig.~\ref{fig:fiducial_pdf_distributions}, approximately $10$ and $6$ (for SKA-1 and SKA-2) DCOs containing NSs will be localised well enough such that, \textit{if} the NS is a pulsar, SKA can unambiguously match it to the radio signal.

If there is more than one pulsar in the region given by the LISA sky localisation, one can compare the measured parameters of the system in LISA and SKA. Both SKA and LISA will measure the orbital frequency to high precision, as well as the time derivative of the frequency and chirp mass to a lesser precision, of each of these systems. Therefore, one could perform a targetted search with SKA that checks the sky location given by LISA and only looking for binary pulsars with orbital frequencies within the errors. If there was \textit{still} more than one possible pulsar one could also check against the chirp mass. In this way, it will be possible to get a joint detection between SKA and LISA even when the sky area implied by the LISA detection contains more than one pulsar.

@@

\textit{Underlying helium star models:} One major weakness is that the \citet{Hurley+2000} fitting formulae for the evolution of helium stars are based on a grid of models from $0.3 \unit{M_{\odot}}$ to $10 \unit{M_{\odot}}$, for a single metallicity ($Z= 0.02$) and thus the formulae have no metallicity dependence and are extrapolated for higher masses. A more detailed set of models in this regime could lead to large changes in the evolution of naked helium stars, a common progenitor of DCOs, and thus affect the detection rate of DCOs.

@@

\textit{Limited metallicity range:} Another limitation of the stellar evolution fitting formulae that COMPAS uses is that they are limited to a metallicity range of $10^{-4} \le Z \le 0.03$ and should not be extrapolated outside this region. However, in the Milky Way (based on the metallicity relation in \citet{Frankel+2018}), the metallicity distribution can extend as far as $10^{-5} \le Z \le 0.06$, with a significant fraction of star formation occurring past $Z = 0.03$. Therefore, for our study we had to reassign any metallicities outside of COMPAS' range. We expect that stellar winds will be reduced to such a degree that they are effectively zero for any metallicity below the minimum. Hence we set any metallicity below the minimum, $Z = 10^{-4}$, equal to the $Z = 10^{-4}$. Similarly, DCO formation is less efficient at high metallicity \citep[e.g.][]{Broekgaarden+2021} and so exploring metallicities above the COMPAS maximum is unlikely to contribute significantly to the observed rate. Therefore, we place any sampled metallicity above the maximum of $Z = 0.03$ uniformly randomly in one of the top 5 highest bins (since using a single bin for many binaries leads to unphysical artifacts).

@@

\textit{Systemic kicks:} We do not include the effect of systemic kicks on the final location of the sources. This would require integrating the orbital evolution of the millions of binaries in our sample and thus was not computationally reasonable to include. We investigated the effect of kicks for a small grid of binaries and found that though they would result in a more spread out distribution within the galaxy (with a smaller concentration in the galaxy centre), the overall distribution of positions would be relatively unchanged and very few sources have strong enough kicks to reach escape velocity for the Milky Way.

\textit{Eccentricity measurement uncertainty:} The method that we use to determine the eccentricity uncertainty is pessimistic as it requires each harmonic to be individually detectable \citep[e.g.][]{Lau+2020}. In reality this may not be necessary depending on the efficacy of matched-filter analysis of LISA data. For an eccentric source to have been detected within the LISA data, several harmonics would already have to have been matched as the same source. This could be done by looking in the same region of the sky for signals with similar chirp masses and distances to the most detectable harmonic in order to find other harmonics that are below the regular detection threshold. This would allow one to refine the measurement of the eccentricity uncertainty significantly by comparing the many different harmonics. Therefore, the eccentricity uncertainty that we calculate in this study is a pessimistic estimate. Smaller eccentricity uncertainties would have two main effects on our results. Firstly, the chirp mass error would decrease slightly in the cases where it is dominated by the eccentricity uncertainty. However, it is mainly dominated by the frequency derivative uncertainty since most sources are essentially stationary and so have extremely small chirps. Secondly, it would improve our ability to distinguish between WDWDs and higher mass DCOs. Until we know more about how LISA will search for eccentric sources, we rely upon our pessimistic estimates.

>>>>

Another discriminator between WDWDs and other DCOs is eccentricity. WDWDs formed in the disc are thought to be formed through isolated binary formation and have little to no eccentricity \citep[e.g.][]{Nelemans+2001}. Therefore, if any system is detected with anything other than one detectable harmonic, this suggests that the system is unlikely to be a WDWD. We find that for a 4(10)-year LISA mission, \BHBHMultipleHarmonicsFourPerc{}(\BHBHMultipleHarmonicsTenPerc{})\% of BHBHs, \BHNSMultipleHarmonicsFourPerc{}(\BHNSMultipleHarmonicsTenPerc{})\% of BHNSs and \NSNSMultipleHarmonicsFourPerc(\NSNSMultipleHarmonicsTenPerc{})\% of NSNSs are detected with multiple harmonics. Both the absolute percentage and the relative improvement with an extended LISA mission is lower for the BHNSs with respect to other DCOs as we find that these BHNSs are less eccentric on average (see Fig.~\ref{fig:fiducial_pdf_distributions} and discussion in Sec.~\ref{sec:fiducial_distributions}).

@@

For separating the BHBH and BHNS population one could do so probabilistically given the properties that are measured, particularly the orbital frequency, mass ratio and eccentricity, since these distributions are fairly different for the two DCO types (see Fig.~\ref{fig:fiducial_pdf_distributions}). This method would pose a challenge however as it would likely only rule out some DCO types as the source of the signal rather than provide strong evidence of the exact type.

Another possible solution would be the existence of electromagnetic counterparts to the gravitational wave signal. In Section~\ref{sec:pulsar_matching} we consider the possibility of detecting a pulsar within a BHNS or NSNS system. This could be used to identify the type of the source, however it is unlikely that a large fraction of the population will contain pulsars that are beaming towards the Earth.

\subsection{Matching LISA detections to pulsars with SKA}\label{sec:pulsar_matching}
Since the vast majority of the LISA detectable population of DCOs will not merge for many years, the main form of electromagnetic counterpart for the this population is pulsars. Therefore, for this section we focus only on BHNSs and NSNSs since no BHBH system will contain a pulsar. The joint detection of a binary pulsar with LISA and SKA would not only help to constrain the parameters of the binary, but also enable investigation of other compact object physics. A pulsar(PSR)+BH can provide stringent tests of theories of gravity, in particular the ``No-hair theorem'' \citep{Keane+2015}. Alternatively, an ultrarelativstic PSR+NS system could be used to measure the neutron star equation of state up to an order of magnitude more accurately than other proposed observational constraints \citep{Kyutoku+2019, Thrane+2020}.

We estimate on average, given the number of detectable pulsars and SKA sky area, each pulsar in SKA occupies a region with an angular resolution of $\sigma_{\theta} < 1.3^\circ$ or $0.7^\circ$ for SKA-1 and SKA-2 respectively (see Appendix~\ref{app:ska_area}). Therefore, any DCOs containing NSs localised by LISA with an angular resolution lower than these values can be unambiguously matched to the radio signal in SKA. By considering Fig.~\ref{fig:ang_res}, approximately $11$ and $6$ (for SKA-1 and SKA-2) will satisfy this constraint.

If there is more than one pulsar in the region given by the LISA sky localisation, one can compare the measured parameters of the system in LISA and SKA. Both SKA and LISA will measure the orbital frequency to high precision, as well as the time derivative of the frequency and chirp mass to a lesser precision, of each of these systems. Therefore, one could perform a targeted search with SKA that checks the sky location given by LISA and only looking for binary pulsars with orbital frequencies within the uncertainties. If there was \textit{still} more than one possible pulsar one could also check against the chirp mass. In this way, we expect it will be possible to get a joint detection between SKA and LISA even when the sky area implied by the LISA detection contains more than one pulsar.

@@

\textit{Underlying helium star models:} One major weakness is that the \citet{Hurley+2000} fitting formulae for the evolution of helium stars are based on a grid of models from $0.3 \unit{M_{\odot}}$ to $10 \unit{M_{\odot}}$, for a single metallicity ($Z= 0.02$) and thus the formulae have no metallicity dependence and are extrapolated for higher masses. A more comprehensive set of models in this regime could lead to large changes in the evolution of naked helium stars, a common progenitor of DCOs, and thus affect the detection rate of DCOs.

@@

\textit{Limited metallicity range:} Another limitation of the stellar evolution fitting formulae that COMPAS uses is that they are limited to a metallicity range of $10^{-4} \le Z \le 0.03$ and should not be extrapolated outside this region. However, in the Milky Way (based on the metallicity relation in \citealt{Frankel+2018}), the metallicity distribution can extend as far as $10^{-5} \le Z \le 0.06$, with a significant fraction of star formation occurring past $Z = 0.03$. Therefore, for our study we had to reassign any metallicities outside of COMPAS' range. We expect that stellar winds will be reduced to such a degree that they are effectively zero for any metallicity below the minimum. Hence we set any metallicity below the minimum, $Z = 10^{-4}$, equal to the $Z = 10^{-4}$. Similarly, DCO formation is less efficient at high metallicity \citep[e.g.][]{Broekgaarden+2021} and so exploring metallicities above the COMPAS maximum is unlikely to contribute significantly to the observed rate. Therefore, we place any sampled metallicity above the maximum of $Z = 0.03$ uniformly randomly in one of the top 5 highest bins that range across $0.01416 < Z < 0.03$ (since using a single metallicity for many binaries leads to unphysical artifacts).

@@

\textit{Systemic kicks:} We do not include the effect of systemic kicks on the final location of the sources. This would require integrating the orbital evolution of the millions of binaries in our sample and thus was not computationally reasonable to include. We investigated the effect of kicks for a small grid of binaries and found that, though they would result in a more spread out distribution (with a smaller concentration in the Galactic centre) and larger heights above the plane, the overall distribution of positions would be relatively unchanged and very few sources have strong enough kicks to reach escape velocity for the Milky Way.

\textit{Eccentricity measurement uncertainty:} The method that we use to determine the eccentricity uncertainty is pessimistic as it requires each harmonic to be individually detectable \citep[e.g.][]{Lau+2020}. In reality this may not be necessary depending on the efficacy of matched-filter analysis of LISA data. For an eccentric source to have been detected within the LISA data, several harmonics would already have to have been matched as originating from the same source. This could be done by looking in the same region of the sky for signals with similar chirp masses and distances to the most detectable harmonic in order to find other harmonics that are below the regular detection threshold. This would allow one to refine the measurement of the eccentricity uncertainty significantly by comparing the many different harmonics. Therefore, the eccentricity uncertainty that we calculate in this study is a pessimistic estimate. Smaller eccentricity uncertainties would have two main effects on our results. Firstly, the chirp mass error would decrease slightly in the cases where it is dominated by the eccentricity uncertainty. However, it is mainly dominated by the frequency derivative uncertainty since most sources are essentially stationary in frequency. Secondly, it would improve our ability to distinguish between WDWDs and higher mass DCOs. Until we know more about how LISA will search for eccentric sources, we rely upon our pessimistic estimates.
